Shaw et al. (2025) Supplementary analysis

### 2025-04-17 scripts/Paper-analysis.Rmd
cleaning up Paper-analysis.Rmd and outputs

```{r}
library(ggplot2)
library(knitr)
library(xtable)
library(dplyr)
library(tidyr)
```

## Compare hard shell genes vs. accessory

@@

Repeated this analysis with one gene representative per family:

@@

# Median of selected k-mers
df = t(data.frame(c(median(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"])$p.value),
        c(median(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"])$p.value),
        c(median(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"])$p.value)))
colnames(df) = c("median", "statistic", "p.value")
print(df)

### All hard shell vs. accessory, one representative per gene family

We subsampled to one representative per gene family and the same number of hard shell and accessory genes, whichever was smaller (PTU-Bac37 was the only one that had more hard shell than accessory genes). 
```{r}
palindromes_k4 = tolower(read.csv('~/Downloads/trieste/k4.txt', header=F)$V1)
palindromes_k6 = tolower(read.csv('~/Downloads/trieste/k6.txt', header=F)$V1)

@@

## Analysis of leading region

Some preliminary statistics about discarded PTUs (for main text).

We discard those where the minimum size is <20kb: E1, E3, E9, E10, E14, E76. We check the maximum size per PTU:

For convenience for other scripts, I save some files with names of plasmids within each PTU.

```
### Sliding window analysis

k=6 
```{r}
# Theoretical expectation for 6-bp palindrome density
theoreticalPalindromeExpectation = function(p){# p is GC content
  return ((p**2 - p + 1/2)**3)
}
d = read.csv('../test_output_window5000_step500.csv', header=F, stringsAsFactors = F)
colnames(d) = c("PTU", "plasmid", "x", "window_size", "palindrome_count", "palindrome_density", "GC_n")

# One line per plasmid
d.reduced.median = d.reduced %>% group_by(PTU, x) %>% 
  summarise(median=median(diff_density),
            n=length(diff_density), 
            median.original=median(palindrome_density),
            GC=median(GC_n))

# Maximum counts
d.reduced.median.plot = d.reduced.median %>% 
  group_by(PTU) %>%
  mutate(max_n = max(n, na.rm = TRUE),  # Calculate max 'n' for each 'PTU'
         normalized_n = n / max_n * 100) %>%  # Normalize 'n' by this max value
  ungroup() %>%  # Remove the grouping
  select(-max_n)  # Optionally remove the max_n column if not needed
# Plot the distribution
p.lines.expectation = ggplot(d.reduced.median.plot, aes(x, median*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome depletion per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.expectation, file='../results/6mer-deviation-expectation.pdf',
       width=10, height=6)
# Plot only first 50kb
ggsave(p.lines.expectation+scale_x_continuous(labels = function(x) x / 1000, limits=c(0,50000)), file='../results/6mer-deviation-expectation-0-50-kb.pdf',
       width=10, height=6)

```
```{r}
# Sliding window for GC content
p.lines.GC = ggplot(d.reduced.median.plot, aes(x, GC/5000 * 100))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("GC content (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.GC, file='../results/GC-content.pdf',
       width=10, height=6)

# And for palindrome density (not expected - actual values)
p.lines.density = ggplot(d.reduced.median.plot, aes(x, median.original*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome density per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.density, file='../results/6mer-palindrome-density.pdf',
       width=10, height=6)

```{r}
diff.PTU.averages = diff.leading.region.average %>% group_by(PTU) %>%
  summarise(mean.GC=mean(diff.GC),
            GC=mean(GC_n[x=="0"]/5000 * 100),
            mean.palindrome=mean(palindrome_density),
            mean.palindrome.diff=mean(diff.density) * 1000,
            n=length(diff.GC))
ggplot(diff.PTU.averages, aes(mean.GC, mean.palindrome.diff))+
  geom_hline(yintercept = 0, linetype='dashed')+
  ggrepel::geom_text_repel(aes(label=PTU))+
    geom_point(aes(size=n), shape=21, colour="black", fill="grey")+
  theme_bw()+
  theme(panel.grid = element_blank())+
  xlab("GC difference between first 5kb\nand plasmid average (%)")+
  ylab("Additional palindrome depletion in leading region (per kb)")

# Calculate actual GC and actual palindrome density
# for a given x-value. I think 5kb is appropriate for leading region to 
# get rid of start effects. 
gc_data_frame = function(position_on_plasmid){
  d.reduced %>% 
  group_by(plasmid, PTU) %>%
  mutate(GC=GC_n[x==position_on_plasmid]/5000 * 100,
         diff.GC=(GC_n[x==position_on_plasmid]-mean(GC_n))/5000 * 100,
         density=palindrome_density[x==position_on_plasmid]*1000,
         density.depletion=diff_density[x==position_on_plasmid]*1000) %>%
  filter(x==position_on_plasmid)
}
d.GC.palindrome = gc_data_frame(5000)
ggplot(d.GC.palindrome, aes(diff.GC, density.depletion))+
  geom_point()

d.GC.palindrome.PTU = d.GC.palindrome %>% group_by(PTU) %>%
  summarise(GC=mean(GC),
            diff.GC=mean(diff.GC),
            density=mean(density),
            density.depletion=mean(density.depletion),
            n=length(window_size))
ggplot(d.GC.palindrome, aes(diff.GC, density.depletion, fill=PTU))+
  geom_point(shape=21, colour="black")+
  facet_wrap(~PTU)
p.gc.correlation = ggplot(d.GC.palindrome, aes(diff.GC, density.depletion))+
  geom_point()+
  theme_bw()+
  xlab("Relative GC content at 5-10kb (%)")+
  ylab("Relative palindrome depletion at 5-10kb (per kb)")+
    xlim(c(-15, 10))+
  ylim(-13, 14)
p.gc.correlation.ptu =  ggplot(d.GC.palindrome.PTU, aes(diff.GC, density.depletion))+
  geom_point(aes(size=n, fill=PTU), shape=21, colour="black")+
  theme_bw()+
  ggrepel::geom_text_repel(aes(label=PTU), size=2, min.segment.length = 0)+
  guides(fill=FALSE)+
    xlab("Relative GC content (%)")+
  ylab("Relative palindrome depletion (per kb)")
pdf('../results/Figure-3-GC-correlation-depletion.pdf', width=8, height=4)
cowplot::plot_grid(p.gc.correlation+ggtitle("(a)"), p.gc.correlation.ptu+ggtitle("(b)"))
dev.off()
cor.test(d.GC.palindrome$diff.GC, d.GC.palindrome$density.depletion, method="spearman")
cor.test(d.GC.palindrome.PTU$diff.GC, d.GC.palindrome.PTU$density.depletion)

summary(lm(density.depletion ~ diff.GC*PTU, data=d.GC.palindrome))

### Averaging over all plasmids

Also do this for GC content.

```{r}
overall.median.GC = d.reduced %>% filter(x<50000) %>%
  group_by(x) %>%
  summarise(y=mean(GC_n),
            n=length(GC_n))
p.overall.average.GC = ggplot(overall.median.GC, aes(x, y/5000 * 100))+
  geom_line()+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 50, linetype='dashed')+
  ylab("GC content (per kb)")+
    scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(axis.text=element_text(colour="black"))
p.overall.average.GC
# Save
ggsave(p.overall.average.GC, file='../results/GC-mean-overall.pdf', 
       width=5, height=2)
```

```{r}
lagging.df = read.csv('~/Downloads/trieste/leading_lagging/NZ_LT994833.1_lagging_10000_k6.csv', header=T)
leading.df = read.csv('~/Downloads/trieste/leading_lagging/NZ_LT994833.1_leading_10000_k6.csv', header=T)
merged.df = merge(leading.df, lagging.df, by="word")
palindromes = tolower(read.csv('~/Downloads/trieste/k6.txt', header=F)$V1)
ggplot(merged.df[which(merged.df$word %in% palindromes),], aes(score.x, score.y))+
  geom_point()
merged.df.palindromes = merged.df[which(merged.df$word %in% palindromes),]
table(merged.df.palindromes$score.x<merged.df.palindromes$score.y)
kmer_targets = tolower(read.csv('~/Downloads/trieste/PTU-LM_targets.txt', header=F)$V1)
ggplot(merged.df[which(merged.df$word %in% kmer_targets),], aes(score.x, score.y))+
  geom_point()
merged.df.targets = merged.df[which(merged.df$word %in% kmer_targets),]
table(merged.df.targets$score.x<merged.df.targets$score.y)

### Palindromes in leading vs. lagging

# Order by average score
merge_scores_counts$kmer = ordered(merge_scores_counts$kmer,
                              levels=scores_average$kmer[order(scores_average$score_avg)])
scores.plot = ggplot(merge_scores_counts, aes(score_avg, kmer))+
    geom_vline(xintercept = 0)+
  geom_bar(stat="identity")+
  geom_vline(xintercept = median(merge_scores_counts$score_avg), linetype='dashed', colour='red')+
  theme_bw()+
  theme(axis.text=element_text(colour="black", size=12),
        axis.title.x=element_text(size=24),
        axis.text.x=element_text(size=18))+
  theme(axis.text.y=element_text(hjust=0))+
  ylab("")+
  xlab("Mean depletion score\n(leading vs. lagging, 10kb)")
wilcox.test(merge_scores_counts$score_avg)
ggsave(scores.plot, file='../results/6mer-palindromes-scores.pdf',
       height=10, width=10)
```

Figure 4: Average depletion of within-range RM targets in the leading vs. lagging regions for conjugative PTUs. 
```{r}

# Order by median score for k=6
order.k.6.scores = results[which(results$k==6), "PTU"][order(results[which(results$k==6), "median.score"])]
results$PTU = ordered(results$PTU,
                      levels=rev(order.k.6.scores))
# Add number of targets to each

p.depletion.targets = ggplot(results, aes(median.score, as.factor(PTU)))+
  geom_vline(xintercept = 0, size=1)+
  geom_bar(stat="identity", position="dodge")+
  facet_wrap(~k, scales="free_x")+
  ylab("")+
  xlab("Median depletion score of within-range RM targets\n(leading vs. lagging, 10kb)")+
   theme_bw()+
  theme(panel.grid = element_blank())+
  theme(axis.text=element_text(colour="black", size=12),
        axis.title.x=element_text(size=24),
        axis.text.x=element_text(size=18))+
  theme(axis.text.y=element_text(hjust=0))
ggsave(p.depletion.targets, file='../results/Figure-4-figure-depletion-within-range-targets.pdf',
       width=10, height=6)

# Is there a correlation between k=5 and k=6 results?
results.df = results %>% select(k, median.score, PTU) %>%
  pivot_wider(
    names_from = k,  # Use values of 'k' as new column names
    values_from = median.score  # Use 'median.score' for the column values
)
```

@@

results.df.norm.average = results.df.long %>% group_by(kmer) %>%
summarise(median=median(score_avg)) %>%
arrange(median)
# Take the top 1% depleted
results.df.norm.average.top.depleted = results.df.norm.average[which(results.df.norm.average$median< quantile(results.df.norm.average$median, probs = c(0.01))),]
results.df.norm.average.top.depleted$palindrome = ifelse(results.df.norm.average.top.depleted$kmer %in% palindromes,
"palindrome",
"non-palindrome")
# Read in all RM targets
all_rm_targets = read.csv('../data/rebase_all_4_6bp_targets_250131.csv', header=F, sep=' ')
all_rm_targets_6bp =all_rm_targets$target[all_rm_targets$length==6]
results.df.norm.average.top.depleted$target = ifelse(results.df.norm.average.top.depleted$kmer %in%
all_rm_targets_6bp,
"target", "non-target")
table(results.df.norm.average.top.depleted$palindrome)
table(results.df.norm.average.top.depleted$target)
chisq.test(matrix(c(7, 34, 865-7, 4096-865-34 ), nrow=2))
chisq.test(matrix(c(4, 37, 60, 4055 ), nrow=2))

>>>>

---
title: "Shaw et al. (2025) Supplementary analysis"
author: "Liam Shaw"
output:
  html_document:
    toc: true
---

### Preliminaries

This script contains analysis and figure generation for Shaw et al. (2025), 'The leading region of many conjugative plasmids is depleted in restriction-modification targets'. For more details, see the associated paper. The code matches the order of appearance of the figures/analysis in the paper and is separated into sections based on the figures. Figure S5 and S6 are not included in this code.

It is intended to be run from inside the `scripts` directory of this repository. Results are saved to to `results` within the repository.

**Note** This relies on large downloaded files and is not intended to work out-of-the-box. For example, Roary gene presence/absence tables on my computer are stored in `/Users/Liam/Downloads/06_plaspan/`.

## Dataset

### Statistics

First, we load libraries.

```{r, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE)
library(ggplot2)
library(knitr)
library(xtable)
library(dplyr)
library(tidyr)
```

We calculate some preliminary statistics about the PTUs from our dataset of 50 PTUs that we analyse (see main text).

We discard those where the minimum size of plasmids in the PTU is <20kb: E1, E3, E9, E10, E14, E76. We check the maximum size per PTU:

For convenience for other scripts, we save some files with names of plasmids within each PTU.

### Figure 1

Figure 1. Summary of the 36 plasmid taxonomic units (PTUs) analysed in this paper.

```{r, message=FALSE}
ptu = read.csv('../data/PTU-pangenome-stats.csv')
for (each_ptu in ptu$PTU){
  gene_pa_file <- list.files(paste0("/Users/Liam/Downloads/06_plaspan/", each_ptu), 
                        pattern = "gene_presence_absence.Rtab$", full.names = TRUE, recursive = TRUE)
  gene_counts = read.csv(gene_pa_file, sep='\t', row.names = 1)
  ptu$median.genes[which(ptu$PTU==each_ptu)] = median(colSums(gene_counts))
  ptu$max.genes[which(ptu$PTU==each_ptu)] = max(colSums(gene_counts))
  ptu$min.genes[which(ptu$PTU==each_ptu)] = min(colSums(gene_counts))
}

# Set up for plotting
ptu_for_plotting = ptu %>% pivot_longer(cols=c("n.core", "n.others"))
# Order by median size (no. of genes)
ptu = ptu[order(ptu$median.genes),]
ptu$PTU = ordered(ptu$PTU, 
                  levels=rev(ptu$PTU))
ptu$leading.region = ifelse(ptu$PTU %in% leading.region.PTUs,
                            "yes", "no")
# Make the figure
pdf('../results/Figure-1-ptu-summary-pangenome-plot.pdf', width=9, height=5)
p.figure1 = ggplot(ptu[which(ptu$median.genes>20),], aes(PTU, median.genes, ymin=min.genes, ymax=max.genes,
                                             fill=leading.region))+
  geom_bar(aes(y=n.core), stat="identity")+
  geom_errorbar()+
  geom_point(shape=21)+
  theme_bw()+
  xlab("")+
  ylab("Number of genes")+
  labs(fill="Leading region analysed")+
  coord_flip()+
  scale_color_brewer(palette="Set2")+
  scale_fill_brewer(palette="Set2")+
  theme(axis.text=element_text(colour="black"))
p.figure1
dev.off()
p.figure1
```

## Sliding window analysis

### Figures S1-S3

We plot results of sliding window analysis over the conjugative plasmids for k=6. This produces Figure

```{r, message=FALSE}
# Theoretical expectation for 6-bp palindrome density
theoreticalPalindromeExpectation = function(p){# p is GC content
  return ((p**2 - p + 1/2)**3)
}
d = read.csv('../test_output_window5000_step500.csv', header=F, stringsAsFactors = F)
colnames(d) = c("PTU", "plasmid", "x", "window_size", "palindrome_count", "palindrome_density", "GC_n")

# One line per plasmid
d.reduced.median = d.reduced %>% group_by(PTU, x) %>% 
  summarise(median=median(diff_density),
            n=length(diff_density), 
            median.original=median(palindrome_density),
            GC=median(GC_n))
# Maximum counts
d.reduced.median.plot = d.reduced.median %>% 
  group_by(PTU) %>%
  mutate(max_n = max(n, na.rm = TRUE),  # Calculate max 'n' for each 'PTU'
         normalized_n = n / max_n * 100) %>%  # Normalize 'n' by this max value
  ungroup() %>%  # Remove the grouping
  select(-max_n)  # Optionally remove the max_n column if not needed

# Sliding window for GC content, S1
p.figureS1 = ggplot(d.reduced.median.plot, aes(x, GC/5000 * 100))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("GC content (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.figureS1, file='../results/Figure-S1-GC-content.pdf',
       width=10, height=6)
p.figureS1

overall.median.GC = d.reduced %>% filter(x<50000) %>%
  group_by(x) %>%
  summarise(y=mean(GC_n),
            n=length(GC_n))
p.figureS1.overall = ggplot(overall.median.GC, aes(x, y/5000 * 100))+
  geom_line()+
  theme_bw()+
  theme(panel.grid = element_blank())+
  geom_hline(yintercept = 50, linetype='dashed')+
  ylab("GC content (per kb)")+
    scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(axis.text=element_text(colour="black"))
p.figureS1.overall
# Save
ggsave(p.figureS1.overall, file='../results/Figure-S1-GC-mean-overall.pdf', 
       width=5, height=2)

# Plot the distribution, S2
p.figureS2 = ggplot(d.reduced.median.plot, aes(x, median*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome depletion per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.figureS2, file='../results/Figure-S2-6mer-deviation-expectation.pdf',
       width=10, height=6)
p.figureS2
```

```{r}
# And for palindrome density (not expected - actual values)
p.figureS3 = ggplot(d.reduced.median.plot, aes(x, median.original*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome density per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.figureS3, file='../results/Figure-S3-6mer-palindrome-density.pdf',
       width=10, height=6)
p.figureS3
```

This is the main panel of the figure.

This is the inset panel which is added into the panel above to create the figure in the paper.

```{r}
diff.PTU.averages = diff.leading.region.average %>% group_by(PTU) %>%
  summarise(mean.GC=mean(diff.GC),
            GC=mean(GC_n[x=="0"]/5000 * 100),
            mean.palindrome=mean(palindrome_density),
            mean.palindrome.diff=mean(diff.density) * 1000,
            n=length(diff.GC))

# Calculate actual GC and actual palindrome density
# for a given x-value. 5kb is appropriate for leading region to 
# get the strongest signal (see Figure 2). 
gc_data_frame = function(position_on_plasmid){
  d.reduced %>% 
  group_by(plasmid, PTU) %>%
  mutate(GC=GC_n[x==position_on_plasmid]/5000 * 100,
         diff.GC=(GC_n[x==position_on_plasmid]-mean(GC_n))/5000 * 100,
         density=palindrome_density[x==position_on_plasmid]*1000,
         density.depletion=diff_density[x==position_on_plasmid]*1000) %>%
  filter(x==position_on_plasmid)
}
d.GC.palindrome = gc_data_frame(5000)

d.GC.palindrome.PTU = d.GC.palindrome %>% group_by(PTU) %>%
  summarise(GC=mean(GC),
            diff.GC=mean(diff.GC),
            density=mean(density),
            density.depletion=mean(density.depletion),
            n=length(window_size))

p.figure3a = ggplot(d.GC.palindrome, aes(diff.GC, density.depletion))+
  geom_point()+
  theme_bw()+
  xlab("Relative GC content at 5-10kb (%)")+
  ylab("Relative palindrome depletion at 5-10kb (per kb)")+
    xlim(c(-15, 10))+
  ylim(-13, 14)
p.figure3b =  ggplot(d.GC.palindrome.PTU, aes(diff.GC, density.depletion))+
  geom_point(aes(size=n, fill=PTU), shape=21, colour="black")+
  theme_bw()+
  ggrepel::geom_text_repel(aes(label=PTU), size=2, min.segment.length = 0)+
  guides(fill=FALSE)+
    xlab("Relative GC content (%)")+
  ylab("Relative palindrome depletion (per kb)")
pdf('../results/Figure-3-GC-correlation-depletion.pdf', width=8, height=4)
p.figure3 = cowplot::plot_grid(p.figure3a+ggtitle("(a)"), p.figure3b+ggtitle("(b)"))
p.figure3
dev.off()
p.figure3
# Calculate correlations
cor.test(d.GC.palindrome$diff.GC, d.GC.palindrome$density.depletion, method="spearman")
cor.test(d.GC.palindrome.PTU$diff.GC, d.GC.palindrome.PTU$density.depletion)
```

## Depletion of specific targets

Figure 4: Average depletion of within-range RM targets in the leading vs. lagging regions for conjugative PTUs.

```{r}

# Order by median score for k=6
order.k.6.scores = results[which(results$k==6), "PTU"][order(results[which(results$k==6), "median.score"])]
results$PTU = ordered(results$PTU,
                      levels=rev(order.k.6.scores))

p.figure4 = ggplot(results, aes(median.score, as.factor(PTU)))+
  geom_vline(xintercept = 0, size=1)+
  geom_bar(stat="identity", position="dodge")+
  facet_wrap(~k, scales="free_x")+
  ylab("")+
  xlab("Median depletion score of within-range RM targets\n(leading vs. lagging, 10kb)")+
   theme_bw()+
  theme(panel.grid = element_blank())+
  theme(axis.text=element_text(colour="black", size=12),
        axis.title.x=element_text(size=24),
        axis.text.x=element_text(size=18))+
  theme(axis.text.y=element_text(hjust=0))
ggsave(p.figure4, file='../results/Figure-4-figure-depletion-within-range-targets.pdf',
       width=10, height=6)
```

### Figure S4

# Order by average score
merge_scores_counts$kmer = ordered(merge_scores_counts$kmer,
                              levels=scores_average$kmer[order(scores_average$score_avg)])
p.figureS4 = ggplot(merge_scores_counts, aes(score_avg, kmer))+
    geom_vline(xintercept = 0)+
  geom_bar(stat="identity")+
  geom_vline(xintercept = median(merge_scores_counts$score_avg), linetype='dashed', colour='red')+
  theme_bw()+
  theme(axis.text=element_text(colour="black", size=12),
        axis.title.x=element_text(size=24),
        axis.text.x=element_text(size=18))+
  theme(axis.text.y=element_text(hjust=0))+
  ylab("")+
  xlab("Mean depletion score\n(leading vs. lagging, 10kb)")
wilcox.test(merge_scores_counts$score_avg)
ggsave(p.figureS4, file='../results/Figure-S4-6mer-palindromes-scores.pdf',
       height=10, width=10)
```

## Other analysis

For more details on the analysis below, see Supplementary Text 1.

@@

We repeat this analysis with one gene representative per family:

@@

# Median of selected k-mers
df = t(data.frame(c(median(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"])$p.value),
        c(median(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"])$p.value),
        c(median(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"])$p.value)))
colnames(df) = c("median", "statistic", "p.value")
print(df)
```

We then subsample to one representative per gene family and the same number of hard shell and accessory genes, whichever is smaller (PTU-Bac37 is the only PTU that had more hard shell than accessory genes).

```{r}
palindromes_k4 = tolower(read.csv('~/Downloads/trieste/k4.txt', header=F)$V1)
palindromes_k6 = tolower(read.csv('~/Downloads/trieste/k6.txt', header=F)$V1)

@@

@@

results.df.norm.average = results.df.long %>% group_by(kmer) %>%
summarise(median=median(score_avg)) %>%
arrange(median)
# Take the top 1% depleted
results.df.norm.average.top.depleted = results.df.norm.average[which(results.df.norm.average$median< quantile(results.df.norm.average$median, probs = c(0.01))),]
results.df.norm.average.top.depleted$palindrome = ifelse(results.df.norm.average.top.depleted$kmer %in% palindromes,
"palindrome",
"non-palindrome")
# Read in all RM targets
all_rm_targets = read.csv('../data/rebase_all_4_6bp_targets_250131.csv', header=F, sep=' ')
all_rm_targets_6bp =all_rm_targets$target[all_rm_targets$length==6]
results.df.norm.average.top.depleted$target = ifelse(results.df.norm.average.top.depleted$kmer %in%
all_rm_targets_6bp,
"target", "non-target")
table(results.df.norm.average.top.depleted$palindrome)
table(results.df.norm.average.top.depleted$target)
chisq.test(matrix(c(7, 34, 865-7, 4096-865-34 ), nrow=2))
chisq.test(matrix(c(4, 37, 60, 4055 ), nrow=2))
```
